RE: Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-05-17 rmd/manuscript.Rmd
edit: text

---
title: "Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems"
output: 
  # word_document:
  #   reference_docx: "word_format.docx"
  pdf_document:
   latex_engine: xelatex
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
csl: https://www.zotero.org/styles/ecosphere
mainfont: Times New Roman
fontsize: 12pt
header-includes:
  \usepackage{lineno, textgreek}
  \linenumbers
---

@@

We simulate trophic community dynamics within branching habitats to investigate the effects of extrinsic controls of FCL. In our study, extrinsic controls include ecosystem geometry [size (number of patches) and complexity (branching probability] and disturbance regime. In addition, our simulation model incorporates established intrinsic drivers of FCL, including the degree of omnivory, productivity level, and species' dispersal ability, allowing us to assess the context dependency on drivers of FCL. We find that ecosystem size and complexity are both positively related to FCL, but the form of this relationship is dependent on other extrinsic and intrinsic factors. Our model is an important step in unifying drivers of FCL across extrinsic and intrinsic drivers and parsing their context-dependency in spatially complex ecosystems.

@@

## State transitions

@@

In the weak and strong omnivory motifs, different state transitions were associated with ecosystem size and complexity. Ecosystem size increased FCL primarily through additive mechanisms, facilitating the state transition of "B" $\rightarrow$ "B + P" (low productivity) or "No species" $\rightarrow$ "B + P" (high productivity), particularly with frequent disturbance (**Figure 5**).

In contrast, insertion (i.e., the insertion of intermediate trophic species) and additive mechanisms were responsible for the effect of ecosystem complexity. With weak omnivory, the transition from "B + P" to "B + C + P" (insertion) was predominant when productivity was low (**Figure 6B**). However, the additive mechanism ("No species" $\rightarrow$ "B + P") took primacy in the high productivity scenario (**Figure 6B**). In the strong omnivory motif, transition patterns depended on disturbance frequency. When disturbance was rare ($p_m = 0.05$), the insertion mechanism ("B + P" $\rightarrow$ "B + C + P") drove the FCL change (**Figure 6C**). As disturbance frequency increased, the primary process switched to an additive mechanism ("B" $\rightarrow$ "B + P" or "No species" $\rightarrow$ "B + P"; **Figure 6C**).

<!-- ## Productivity effect -->

<!-- High productivity strengthened the relationship between FCL and ecosystem geometry. In the chain motif, productivity greatly increased the overall FCL (**Figure S7**), thereby shifting the upper bound of FCL (**Figures 2 and 3**). In contrast, increased productivity decreased FCL with weak and strong omnivory (**Figures 2, 3, and S7**). This negative effect of increased productivity was apparent in smaller and/or simpler ecosystems, steepening the slopes between FCL and ecosystem geometry (**Figures 2 and 3**). -->

<!-- State transitions associated with increased productivity depended on the motif structure. In the chain motif, the transition from "B + C" to "B + C + P" was dominant regardless of disturbance frequency (**Figure S8**). The introduction of omnivory allowed the transition between "B + P" and "B + C + P". With weak omnivory, the transition of "B + C + P" $\rightarrow$ "B + P" was prevalent and shortened FCL across disturbance regimes (**Figure S8**). However, this transition was marginal in the strong omnivory motif. Instead, productivity significantly increased the proportion of the "No species" state (**Figure S8**) in disturbed scenarios, especially in small and/or simple ecosystems (**Figures S5 and S6**). -->

<!-- ## Dispersal effect -->

<!-- Dispersal distance had a positive influence on FCL regardless of simulation scenarios (**Figure S7**). However, this direct effect was weak in comparison to other factors. -->

<!-- State transitions occurred through an additive (chain) or insertion mechanism (omnivory). In the chain motif, the transitions were either "B" $\rightarrow$ "B + C" or "B + C" $\rightarrow$ "B + C+ P", depending on disturbance frequency (**Figure S9**). In the omnivory motifs, the predominant transition was from "B + P" to "B + C + P" (**Figure S9**). -->

@@

```{r}
source(here::here("code/table_param.R"))

knitr::kable(df_value %>%
               dplyr::select(-"One-patch") %>% 
               rename(Value = "Main"),
             escape = FALSE,
             booktabs = TRUE,
             caption = "Parameter values used in the main simulation") %>% 
  kableExtra::kable_styling(latex_options = 'HOLD_position')
```

@@

**Figure 1** (A) Example of simulated branching networks. Node color is reflective of the disturbance value calculated in network generation. (B) Density distributions of disturbance intensity (i.e., proportional reduction of a population). Three networks of 50 patches were produced with varied branching probabilities, which were shown in different colors. The vertical line indicates the mean disturbance intensity at headwaters ($\mu_m = 0.8$). (C) Schematic diagram of trophic interactions in a three-species community. $B$ refers to a basal species whose population level is controlled by a density-dependent function. The consumer species (or intraguild prey), $C$, consumes $B$ as a resource. The predator (or intraguild predator), $P$, is able to consume both $B$ and $C$. The structure and strength of these trophic interactions determined by functional responses. On the far left, a simple linear food chain exists when the predator does not consume the basal species. Moving to the right, basal species' biomass is converted to predator biomass at higher efficiencies with greater attack rates. (D) Proportional contribution of basal species $B$ to predator $P$ under different omnivory scenarios. The contribution changes with the relative density of $B$ to $C$ ($C=50$ in this example), and the rate of change is determined by attack rate and handling time.

**Figure 2** Relationships between food chain length and ecosystem size with high disturbance intensity $\mu_m = 0.8$. Lines indicate loess curves, and shades are associated 95% confidence intervals. Rows distinguish food web motifs, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities.

**Figure 3** Relationships between food chain length and ecosystem complexity with high disturbance intensity $\mu_m = 0.8$. Lines indicate loess curves, and shades are associated 95% confidence intervals. Rows distinguish food web motifs, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities.

**Figure 4** Relationships between patch state occupancy and ecosystem size with high disturbance intensity $\mu_m = 0.8$. Lines indicate loess curves, and shades are associated 95% confidence intervals. Panels indicate different food web motifs. In each panel, rows distinguish patch state, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities.

**Figure 5** Relationships between patch state occupancy and ecosystem complexity with high disturbance intensity $\mu_m = 0.8$. Lines indicate loess curves, and shades are associated 95% confidence intervals. Panels indicate different food web motifs. In each panel, rows distinguish patch state, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities.

>>>>

---
title: "Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems"
output: 
  word_document:
    reference_docx: "word_format.docx"
  # pdf_document:
  #  latex_engine: xelatex
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
csl: https://www.zotero.org/styles/ecosphere
mainfont: Times New Roman
fontsize: 12pt
header-includes:
  \usepackage{lineno, textgreek}
  \linenumbers
---

@@

We simulate trophic community dynamics within branching habitats to investigate the effects of extrinsic controls of FCL. In our study, extrinsic controls include "ecosystem geometry" [size (number of patches) and complexity (branching probability] and disturbance regime. In addition, our simulation model incorporates established intrinsic drivers of FCL, including the degree of omnivory, productivity level, and species' dispersal ability, allowing us to assess the context dependency on drivers of FCL. We find that ecosystem size and complexity are both positively related to FCL, but the form of this relationship is dependent on other extrinsic and intrinsic factors. Our model is an important step in unifying drivers of FCL across extrinsic and intrinsic drivers and parsing their context-dependency in spatially complex ecosystems.

@@

@@

In the weak and strong omnivory motifs, different state transitions were associated with ecosystem size and complexity. Ecosystem size increased FCL primarily through additive mechanisms, facilitating the state transition of "B" $\rightarrow$ "B + P" (low productivity) or "No species" $\rightarrow$ "B + P" (high productivity), particularly with frequent disturbance (**Figure 5**). In contrast, insertion (i.e., the insertion of intermediate trophic species) and additive mechanisms were responsible for the effect of ecosystem complexity. With weak omnivory, the transition from "B + P" to "B + C + P" (insertion) was predominant when productivity was low (**Figure 6B**). However, the additive mechanism ("No species" $\rightarrow$ "B + P") took primacy in the high productivity scenario (**Figure 6B**). In the strong omnivory motif, transition patterns depended on disturbance frequency. When disturbance was rare ($p_m = 0.05$), the insertion mechanism ("B + P" $\rightarrow$ "B + C + P") drove the FCL change (**Figure 6C**). As disturbance frequency increased, the primary process switched to an additive mechanism ("B" $\rightarrow$ "B + P" or "No species" $\rightarrow$ "B + P"; **Figure 6C**).

## Direct effects of productivity and dispersal

In the chain motif, productivity greatly increased the overall FCL (**Figure S5**), thereby shifting the upper bound of FCL (**Figures 3 and 4**). In contrast, increased productivity decreased FCL with weak and strong omnivory (**Figures 3, 4, and S5**). This negative effect of increased productivity was apparent in smaller and/or simpler ecosystems, steepening the slopes between FCL and ecosystem geometry (**Figures 3 and 4**). State transitions associated with increased productivity depended on the motif structure. In the chain motif, the transition from "B + C" to "B + C + P" was dominant regardless of disturbance frequency (**Figure S8**). The introduction of omnivory allowed the transition between "B + P" and "B + C + P". With weak omnivory, the transition of "B + C + P" $\rightarrow$ "B + P" was prevalent and shortened FCL across disturbance regimes (**Figure S8**). However, this transition was marginal in the strong omnivory motif. Instead, productivity significantly increased the proportion of the "No species" state (**Figure S8**) in disturbed scenarios, especially in small and/or simple ecosystems (**Figures S5 and S6**).

Dispersal distance had a positive influence on FCL regardless of simulation scenarios (**Figure S5**). State transitions occurred through an additive (chain) or insertion mechanism (omnivory). In the chain motif, the transitions were either "B" $\rightarrow$ "B + C" or "B + C" $\rightarrow$ "B + C+ P", depending on disturbance frequency (**Figure S9**). In the omnivory motifs, the predominant transition was from "B + P" to "B + C + P" (**Figure S9**). However, this direct effect was weak in comparison to other factors with no qualitative alteration to the relationships between FCL and ecosystem geometry (**Figures 3 and 4**).

@@

<!-- ```{r} -->
<!-- source(here::here("code/table_param.R")) -->

<!-- knitr::kable(df_value %>% -->
<!--                dplyr::select(-"One-patch") %>%  -->
<!--                rename(Value = "Main"), -->
<!--              escape = FALSE, -->
<!--              booktabs = TRUE, -->
<!--              caption = "Parameter values used in the main simulation") %>%  -->
<!--   kableExtra::kable_styling(latex_options = 'HOLD_position') -->
<!-- ``` -->

@@

**Figure 1** (A) Example of simulated branching networks. Node color is reflective of the disturbance value calculated in network generation. (B) Density distributions of disturbance intensity (i.e., proportional reduction of a population). Three networks of 50 patches were produced with varied branching probabilities, which were shown in different colors. The vertical line indicates the mean disturbance intensity at headwaters ($\mu_m = 0.8$). (C) Schematic diagram of trophic interactions in a three-species community. $B$ refers to a basal species whose population level is controlled by a density-dependent function. The consumer species (or intraguild prey), $C$, consumes $B$ as a resource. The predator (or intraguild predator), $P$, is able to consume both $B$ and $C$. The structure and strength of these trophic interactions determined by functional responses. On the far left, a simple linear food chain exists when the predator does not consume the basal species. Moving to the right, basal species' biomass is converted to predator biomass at higher efficiencies with greater attack rates. Arrow colors schematically indicate the relative reliance of P on B and C, with black indicating trophic specialization. (D) Proportional contribution of basal species $B$ to predator $P$ under different omnivory scenarios. The contribution changes with the relative density of $B$ to $C$ ($C=50$ in this example), and the rate of change is determined by attack rate and handling time.

**Figure 2** Summary of our simulation results. Black text in each node indicates the potential factors that may influence the association between food chain length and ecosystem geometry, and black text on arrows indicates simulated scenarios for a given factor. Blue text denotes the resultant association between food chain length and ecosystem geometry. Panel (A) indicates the effect of ecosystem size. Panel (B) indicates the effect of ecosystem complexity.

**Figure 3** Relationships between food chain length and ecosystem size with high disturbance intensity $\mu_m = 0.8$. Lines in each panel indicate loess curves fitted to $500$ simulation replicates, and shades are associated 95% confidence intervals. Rows distinguish food web motifs, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities. In this figure, ecosystem complexity values for individual simulation replicates were drawn randomly from a uniform distribution $P_b \sim \text{Unif}(0.01, 0.99)$.

**Figure 4** Relationships between food chain length and ecosystem complexity with high disturbance intensity $\mu_m = 0.8$. Lines in each panel indicate loess curves fitted to $500$ simulation replicates, and shades are associated 95% confidence intervals. Rows distinguish food web motifs, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities. In this figure, ecosystem size values for individual simulation replicates were drawn randomly from a uniform distribution $N_p \sim \text{Unif}(10, 150)$.

**Figure 5** Relationships between the proportion of each patch state $\bar{\psi}$ (colors indicate different patch state) and ecosystem size with high disturbance intensity $\mu_m = 0.8$. The proportion of each patch state $\bar{\psi}$ is averaged over time for each simulation replicate. Lines indicate loess curves fitted to $500$ simulation replicates, and shades are associated 95% confidence intervals. Panels indicate different food web motifs. In each panel, rows distinguish disturbance probability $p_m$, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities.

**Figure 6** Relationships between the proportion of each patch state $\bar{\psi}$ (colors indicate different patch state) and ecosystem complexity with high disturbance intensity $\mu_m = 0.8$. The proportion of each patch state $\bar{\psi}$ is averaged over time for each simulation replicate. Lines indicate loess curves fitted to $500$ simulation replicates, and shades are associated 95% confidence intervals. Panels indicate different food web motifs. In each panel, rows distinguish disturbance probability $p_m$, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities.

### 2023-05-17 rmd/reply2.Rmd
edit: text

> This is a great idea! We included a new hierarchical tree figure to show how different factors determine the FCL response to ecosystem size/complexity (Figure 2 in the revised MS).
Also, we revised the figure captions to better guide readers.

@@

> We greatly appreciate your input. We have made improvements to the figure captions in order to enhance reader comprehension regarding the meaning of the figures.

> Specifically, we have carefully edited the captions to provide a clearer understanding of the y-axis by replacing the confusing term "occupancy" with the more explicit phrase "proportion of patch state."

@@

> Thank you, please let us clarify.
We defined FCL as trophic position, not trophic level.
Thus, the FCL can take continuous values.
With omnivory, the max is greater than 2. This occurs when P and C coexist.
P's trophic level depends on the relative abundance of B and C (see equations 12 & 13).
Thus, with omnivory, the FCL is unavoidably shorter than the chain motif even if the same set of species coexist.

@@

> Thank you for your input. "*Looking at the figure 1 c (and from experience) the expectation is that as omnivory increases the FCL decreases.*" -- The response of FCL to increasing omnivory can be more complex.
Weak omnivory can cause similar or even lower FCL than strong omnivory because the intraguild predator can have a greater predation pressure on intraguild prey (due to higher attack rate to C).
For the definition of FCL in our study, please see equations 12 & 13 -- it accounts for relative availability of B and C in a given habitat patch.
Please see also our response to your earlier comments.

@@

> Thank you, we have revised figure captions.
Also, we clarified the values of ecosystem size or complexity in the main figures (both are randomly drawn from uniform distributions).

@@

@@

> We appreciate your understanding regarding the two levels of productivity in our model

> However, we feel that placing excessive emphasis on this particular aspect might further complicate an already complex content, as you astutely pointed out earlier during this round of review.

@@

> Thank you. In the previous version of the manuscript, we have introduced "ecosystem geometry" as the term including ecosystem size and complexity in the Introduction. However, it was probably difficult to find. We have re-worded so the readers can find the information more easily.

>>>>

> This is a great idea! We included a new hierarchical tree figure to show how different factors determine the FCL response to ecosystem size/complexity (Figure 2 in the revised MS).
We revised the results section to reflect the addition of this new figure.
Also, we editted figure captions as appropriate.

@@

> We greatly appreciate your input. We have made improvements to the figure captions in order to enhance reader comprehension regarding the meaning of the figures.
Specifically, we have carefully edited the captions to provide a clearer understanding of the y-axis by replacing the confusing term "occupancy" with the more explicit phrase "proportion of patch state." We also changed the presentation so the readers can understand what transitions happened (color codes changed).

@@

> Thank you, please let us clarify.
We defined FCL as trophic position, not trophic level.
Thus, the FCL can take continuous values.
With omnivory, the max is greater than 2. This occurs when P and C coexist.
P's trophic position depends on the relative abundance of B and C (see equations 12 & 13).
Thus, with omnivory, the FCL is unavoidably shorter than the chain motif even if the same set of species coexist.

@@

> Thank you for your input. "*Looking at the figure 1 c (and from experience) the expectation is that as omnivory increases the FCL decreases.*" -- After some thought, we realized that the response of FCL to increasing omnivory can be more complex.
Weak omnivory can cause similar or even lower FCL than strong omnivory because the intraguild predator can have a greater predation pressure on intraguild prey (due to higher attack rate to C).
We have rephrased the corresponding paragraph in the Introduction to reflect this.

> For the definition of FCL in our study, please see equations 12 & 13 -- it accounts for relative availability of B and C in a given habitat patch.
Please see also our response to your earlier comments.

@@

> Thank you, we have revised figure captions throughout the MS.
We clarified the values of ecosystem size or complexity in the main figures (both are randomly drawn from uniform distributions).

@@

> Also, we thought those state transtion figures are better placed in the main text so the readers can see the associated text more easily.
We have moved those figures to the main text.

@@

> We appreciate your understanding regarding the two levels of productivity in our model.
In the meantime, we feel that placing excessive emphasis on this particular aspect might further complicate an already complex content, as you astutely pointed out earlier during this round of review.

@@

> Thank you.
In the previous version of the manuscript, we have introduced "ecosystem geometry" as the term including ecosystem size and complexity in the Introduction.
However, it was probably difficult to find.
We have editted text so the readers can find the information more easily.

### 2023-05-17 rmd/suppl_info.Rmd
edit: text

```{r fig3, fig.cap="Relationships between food chain length and ecosystem size with low disturbance intensity $\\mu_m = 0.2$. Lines indicate loess curves, and shades are associated 95% confidence intervals. Rows distinguish food web motifs, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities."}
knitr::include_graphics(here::here("figure/figure_si_n_patch.pdf"))
```

@@

```{r fig4, fig.cap="Relationships between food chain length and ecosystem complexity with low disturbance intensity $\\mu_m = 0.2$. Lines indicate loess curves, and shades are associated 95% confidence intervals. Rows distinguish food web motifs, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities."}
knitr::include_graphics(here::here("figure/figure_si_p_branch.pdf"))
```

@@

```{r fig7, fig.cap="Effects of productivity (left column) and dispersal (right column) on food chain length grouped by disturbance frequency ($p_m$; x-axis). Colors distinguish productivity or dispersal scenarios, and rows identify food web motifs. Disturbance intensity $\\mu_m = 0.8$."}
knitr::include_graphics(here::here("figure/figure_si_fcl_r_disp.pdf"))
```

@@

```{r fig8, fig.cap="Effects of productivity on state transitions with high disturbance intensity $\\mu_m = 0.8$. Colors distinguish productivity scenarios, columns represent different disturbance probability $p_m$, and rows identify food web motifs."}
knitr::include_graphics(here::here("figure/figure_si_state_r.pdf"))
```

@@

```{r fig9, fig.cap="Effects of dispersal on state transitions with high disturbance intensity $\\mu_m = 0.8$. Colors distinguish dispersal scenarios, columns represent different disturbance probability $p_m$, and rows identify food web motifs."}
knitr::include_graphics(here::here("figure/figure_si_state_disp.pdf"))
```

@@

```{r fig10, fig.cap="Ecosystem size and complexity dictate total carrying capacity (= $\\sum_x K_x$) and mean distance between a given pair of habitat patches within a network."}
knitr::include_graphics(here::here("figure/figure_si_k_distance.pdf"))
```

>>>>

```{r fig3, fig.cap="Relationships between food chain length and ecosystem size with low disturbance intensity $\\mu_m = 0.2$.  Lines in each panel indicate loess curves fitted to $500$ simulation replicates, and shades are associated 95% confidence intervals. Rows distinguish food web motifs, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities. In this figure, ecosystem complexity values for individual simulation replicates were drawn randomly from a uniform distribution $P_b \\sim \\text{Unif}(0.01, 0.99)$."}
knitr::include_graphics(here::here("figure/figure_si_n_patch.pdf"))
```

@@

```{r fig4, fig.cap="Relationships between food chain length and ecosystem complexity with low disturbance intensity $\\mu_m = 0.2$. Lines in each panel indicate loess curves fitted to $500$ simulation replicates, and shades are associated 95% confidence intervals. Rows distinguish food web motifs, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities. In this figure, ecosystem complexity values for individual simulation replicates were drawn randomly from a uniform distribution $N_p \\sim \\text{Unif}(10, 150)$."}
knitr::include_graphics(here::here("figure/figure_si_p_branch.pdf"))
```

@@

```{r fig5, fig.cap="Effects of productivity (left column) and dispersal (right column) on food chain length grouped by disturbance frequency ($p_m$; x-axis). Box plots show median (horizontal center line) and quartiles (box limits) of $500$ simulation replicates with whiskers extending up to the most extreme data points that are within ±1.5 interquartile range. Dots are outliers extending beyond ±1.5 interquartile range. Colors distinguish productivity or dispersal scenarios, and rows identify food web motifs. Disturbance intensity $\\mu_m = 0.8$."}
knitr::include_graphics(here::here("figure/figure_si_fcl_r_disp.pdf"))
```

@@

```{r fig6, fig.cap="Effects of productivity on state transitions with high disturbance intensity $\\mu_m = 0.8$. The proportion of each patch state $\\bar{\\psi}$ is averaged over time for each simulation replicate. Box plots show median (horizontal center line) and quartiles (box limits) of $500$ simulation replicates with whiskers extending up to the most extreme data points that are within ±1.5 interquartile range. Dots are outliers extending beyond ±1.5 interquartile range. Colors distinguish productivity scenarios, columns represent different disturbance probability $p_m$, and rows identify food web motifs."}
knitr::include_graphics(here::here("figure/figure_si_state_r.pdf"))
```

@@

```{r fig7, fig.cap="Effects of dispersal on state transitions with high disturbance intensity $\\mu_m = 0.8$. The proportion of each patch state $\\bar{\\psi}$ is averaged over time for each simulation replicate. Box plots show median (horizontal center line) and quartiles (box limits) of $500$ simulation replicates with whiskers extending up to the most extreme data points that are within ±1.5 interquartile range. Dots are outliers extending beyond ±1.5 interquartile range. Colors distinguish dispersal scenarios, columns represent different disturbance probability $p_m$, and rows identify food web motifs."}
knitr::include_graphics(here::here("figure/figure_si_state_disp.pdf"))
```

@@

```{r fig8, fig.cap="Ecosystem size and complexity dictate total carrying capacity (= $\\sum_x K_x$) and mean distance between a given pair of habitat patches within a network."}
knitr::include_graphics(here::here("figure/figure_si_k_distance.pdf"))
```
